---
title: "Statistical Rethinking Lecture 02"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

<b>01. Introduction</b>

Globe tossing data set:
L W L L W W W L W W

How should we use the sample?
How to produce a summary?
How to represent uncertatiny? More data is better obviously.

Use Bayesian data analysis:
For each possible explanation of the data,
Count all the ways data can happen.
Explanations with more ways to produce the data are more plausible.

This is too complicated to start with applying to globe problem,
each proportion is a possible explanation and there are infinite
number of real numbers between 0 and 1.

Start with something easier.

<b>02. Garden of Forking Data</b>

4 marbles in a bag.

Possible contents:
(1) 0 b, 4 w (0) # Forbidden by logic
(2) 1 b, 3 w (3 ways to produce)
(3) 2 b, 2 w (8 ways to produce)
(4) 3 b, 1 w (9 ways to produce)
(5) 4 b, 0 w (0) # Forbidden by logic

Observe B, W, B

Bayes method:
Take each possible explanation and count all the ways that can arise given that assumption. Gonna assume (2) as reality. (<b>Figure 2.2</b>: Draw out all possible events from each sample)

<b>03. Counts to plausibility</b>
Unglamorous basis of applied probability:
<i>Things that can happen more ways are plausible</i>

p = proportion of blue marble in the bag

(1) 0 b, 4 w (p=0) (0) # Forbidden by logic
(2) 1 b, 3 w (p=.25) (3 ways to produce)
(3) 2 b, 2 w (p=.5) (8 ways to produce)
(4) 3 b, 1 w (p=.75) (9 ways to produce)
(5) 4 b, 0 w (p=1) (0) # Forbidden by logic

plausibility:
0
0.15
0.40
0.45
0

<b>Plausibility</b>:
```{r}
ways <- c(3, 8, 9)
ways/sum(ways)
```

<b>04. Updating</b>

When new data arrives we don't have to redo the whole calculation, just take old answer and update.

i.e we draw another marble and it is B.

Instead of redrawing garden, we can multiply the previous counts.

<b>Bayesian updating</b>:

RULES
1. State a causual model for how the observations arise, given each possible explanation

2. Count ways data could arrise for each explanation

3. Relative plausibility is relative from rule 2

<b>04. Forking Globe Tossing</b>

Desnity: Relative plausibility

```{r}
plot (0:10 * 10:0 * 10:0 * 10:0)
```

Curve gets narrow and taller, area under the curve is fixed to 1. As it gets concentrated it gets taller because the area needs to be the same.

(1) No minimum sample size in this business. How we explore implications of the model before adding the evidence. You don't learn a lot from one sample, one data point is worth updating because most explanations are still in play.

(2) Shape of posterior distribution embodies sample size. In other frameworks, sample size has a magical quality, but no comparable role in Bayesian analysis. Sample size does matter though, more data = curve gets narrow and taller.

(3) No point estimate. The estimate is the curve. On a curve you can take a lot of points, like mode or mean to summarize a curve.The distribution is the estimate, always use the entire distribution. Summary is always teh last step.

(4) Intervals do not have a strong role to play in Bayesian inference. You can report any interval you want, all it does communicates the shape of the posterior distribution. "95% is obvious superstition. Nothing magic happens at the boundary." That an arbitrary interval contains an arbitray value  is not meaningful. <b>Use the whole distribution.</b>