---
title: "Statistical Rethinking Homework Week 01"
output: html_notebook
---


**Suppose the globe tossing data had turned out to be 4 water and 11 land. Construct the posterior distibution using the grid approximation. Use the same flat prior as in the book.**
```{r}
p_grid <- seq(from=0, to=1, length.out=1000)
prob_p <- rep(1, 1000)
prob_data <- dbinom(4, size=15, prob=p_grid)
posterior <- prob_data * prob_p
posterior <- posterior / sum(posterior)
set.seed(100)
samples <- sample(p_grid, prob=posterior, size=1e4, replace=TRUE)
plot(samples)
```

**Now suppose the data are 4 water and 2 land. Compute the posterior again, but this time use a prior that is zero below p = 0.5 and a constant above p = 0.5. This corresponds to prior information that a majority of the Earth's surface is water.**
```{r}
p_grid <- seq(from=0, to=1, length.out=1000)
prior <- c(rep(0, 500), rep(1, 500))
prob_data <- dbinom(4, size=6, prob=p_grid)
posterior <- prob_data * prior
posterior <- posterior / sum(posterior)
set.seed(100)
samples2 <- sample(p_grid, prob=posterior, size=1e4, replace=TRUE)
plot(samples2)
```

Note: Impossible value less than 0.5 ruled out, the model piles up at 0.5 and higher.

**For the posterior distribution from 2, compute 89% percentile and HPDI intervals. Compare the widths of these intervals. Which is wider? Why? If you had only the information in the interval, what might you understand about the shape of the posterior distribution?**
```{r}
library(rethinking)
set.seed(100)
PI(samples2)
HPDI(samples2)
```

PI (percentile interval is wider). The lower bound is above 0.5 (0.52) and the uppper bound is (0.88). HPDI is more narrow, includes the higher posterior (figure 3.3 in the book). Boundaries of intervals aren't really informative though, nothing happens at the boundary.

**Suppose there is bias in sampling so that Land is more likely than Water to be recorded. Specifically assume that 1-in-5 (20%) of Water samples are accidentally recorded instead as "Land". First, write a generative simulation of this sampling process. Assuming the true proportion of Water is 0.70, what proporition does your simulation tend to produce instead? Second using a simulated sample of 20 tosses, compute the unbiased posterior distribution of the true proportion of water.**
```{r}
set.seed(100)
W <- rbinom(N, size=20, prob=0.7*0.8)
mean(W/20)
```

Biased is .56

```{r}
set.seed(100)
W <- rbinom(1, size=20, prob=0.7*0.8)
p_grid <- seq(from=0, to=1, len=100)
pr_p <- dbeta(p_grid, 1, 1)
prW <- dbinom(W, 20, p_grid*0.8)
post <-prW*pr_p
post_bias <- dbinom(W, 20, p_grid)
plot(p_grid, post, type="l", lwd=4, 
     xlab="proportion water", ylab="plausibility")
lines(p_grid, post_bias, col=2, lwd=4)
```



